---
title: "Escapement goals eval"
author: "Erik Schoen"
date: "10/17/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r load packages, warning = F, message = F}
library(dplyr)
library(tidyr)
library(lubridate)
```

# Prepare daily escapement data

+ read in daily data from [Alaska Department of Fish and Game. 2017. Daily salmon escapement counts from the OceanAK database, Alaska, 1921-2017. Knowledge Network for Biocomplexity. doi:10.5063/F1TX3CKH.](doi:10.5063/F1TX3CKH)

```{r read in daily escapement data}
esc_daily <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")

head(esc_daily)
```

+ filter for Bristol Bay sockeye
+ summarize by river and year, calculating total count

```{r clean and summarize escapement data by location and year}
levels(esc_daily$SASAP.Region)
esc_annual <- esc_daily %>%
  filter(SASAP.Region == "Bristol Bay" & Species == "Sockeye") %>%
  mutate(Date = ymd(sampleDate)) %>%
  mutate(Year = year(Date)) %>%
  group_by(Location, Year) %>%
  summarize(Esc_Annual = sum(DailyCount, na.rm = T))

head(esc_annual)
```


# Prepare escapement goals data

```{r read in escapement goals data}
esc_goals <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/knb.92014.1", stringsAsFactors = F)

head(esc_goals)
```
```{r clean up escapement goals}
esc_goals_clean <- esc_goals %>%
  filter(Region == "Bristol Bay" & Species == "Sockeye") %>%
  filter(Lower != "eliminated") %>%
  select(Region, Location = System, Lower, Upper, Type, Initial.Year) %>%
  mutate(Lower = as.numeric(Lower),
         Initial.Year = as.numeric(Initial.Year))

head(esc_goals_clean)
```


# Evaluate escapement goals

```{r join escapement data to goals}
esc_eval <- left_join(esc_annual, esc_goals_clean, by = "Location")

# Too many rows because there are 2 escapement goals for the Nushagak
# filter out the OEG rows

esc_eval2 <- esc_eval %>%
  filter(Type != "OEG")

# Evaluate whether the escapement goals were met
esc_eval2 <- esc_eval %>%
  mutate(Goal_Met = ifelse(Esc_Annual >= Lower & Esc_Annual <= Upper, T, F))
esc_eval2
```

